---
title: "Homework 14"
format: html
---

```{r}
# For reproducible results
set.seed(1)
# Loading Packages
library(ggplot2)
library(dplyr)
library(psych)        # Bartlett's test
library(randomForest) # Random forest
library(palmerpenguins)

# Data
data(penguins)

head(penguins)
str(penguins)
```
#PCA
#The dataset describes a collection of biological and environmental observations for three different species of penguins
```{r}

# Create a figure showing PC1 and PC2 with 95% CI ellipses
## Do you see good separation of your groups on either axis? 

# Select only numeric columns
penguins_num <- penguins %>%
  select(where(is.numeric)) %>%
  na.omit()

# Species column
penguins_species <- penguins$species

# Correlation matrix
round(cor(penguins_num), 2)
```
```{r}
bart <- cortest.bartlett(cor(penguins_num), n = nrow(penguins_num))
bart
```

```{r}
pca <- prcomp(penguins_num, center = TRUE, scale. = TRUE)
summary(pca)
```
```{r}
eig <- pca$sdev^2
pve <- eig / sum(eig)

pca_var_table <- data.frame(
  PC = paste0("PC", 1:length(eig)),
  Eigenvalue = round(eig, 3),
  PVE = round(pve, 3),
  CumPVE = round(cumsum(pve), 3)
)
pca_var_table
```
```{r}
plot(eig, type = "b", pch = 19,
     xlab = "Principal component",
     ylab = "Eigenvalue",
     main = "Scree plot (penguins PCA)")
```
```{r}
# test
test <- function(p) sapply(1:p, function(k) sum(1/(k:p)) / p)
te <- test(ncol(penguins_num))

retain <- data.frame(
  PC = paste0("PC", 1:length(pve)),
  ObservedPVE = round(pve, 3),
  test = round(te, 3),
  Keep = pve > te
)
retain
```

```{r}
head(pca$x)
```
```{r}

penguins_clean <- penguins %>%
  select(species, where(is.numeric)) %>%
  na.omit()
penguins_num <- penguins_clean %>%
  select(-species)

# Species vector
penguins_species <- penguins_clean$species

# PCA
pca <- prcomp(penguins_num, scale. = TRUE)

# PCA scores
scores <- as.data.frame(pca$x)

# Add species
scores$Species <- penguins_species
head(pca$x)
pca$rotation
```
```{r}
scores <- as.data.frame(pca$x)
scores$Species <- penguins_species

plt <- ggplot(scores, aes(PC1, PC2, color = Species)) +
  geom_point(size = 2.6, alpha = 0.85) +
  stat_ellipse() +
  theme_minimal() +
  labs(
    title = "PCA on Penguins",
    subtitle = "PCA is unsupervised; species used only for coloring"
  )

plt
```
#There is good separation of the Gentoo group from the other two species along the PC1 axis. However, the Adelie and Chinstrap groups show significant overlap on both axes, though they are somewhat distinguishable.
```{r}
man <- manova(cbind(PC1, PC2) ~ Species, data = scores)
summary(man, test = "Pillai")
```
```{r}
set.seed(60)
id_train <- sample(seq_len(nrow(penguins)), size = 0.7 * nrow(penguins))
train <- penguins[id_train, ]
test  <- penguins[-id_train, ]

```
```{r}
set.seed(100)

train_ind <- sample(
  1:nrow(penguins_clean),
  0.7 * nrow(penguins_clean)
)

train <- penguins_clean[train_ind, ]
test  <- penguins_clean[-train_ind, ]
print(head(train))

#Random forest model
rf <- randomForest(
  species ~ .,
  data = train,
  ntree = 500,
  mtry = 2,
  importance = TRUE
)

# Print model
rf
```
#The Out-of-Bag (OOB) estimate of error rate is 2.93%.
#Based on the confusion matrix for the test data, the model was highly accurate, with an overall accuracy of 0.9902913 (approx. 99%)
```{r}
# Predictions
pred <- predict(rf, newdata = test)
table(
  Observed = test$species,
  Predicted = pred
)
```
```{r}
acc <- mean(pred == test$species)
acc
```

```{r}
plot(rf, main = "Random forest OOB error vs number of trees")
```
```{r}
importance(rf)
```
```{r}
varImpPlot(rf)
```
```{r}
#Based on GINI, the most important variables were bill_length_mm and flipper_length_mm.  

#The variable year was consistently the least important across both Random Forest metrics.
```